---
title: "Individual Project"
author: "Nagaprasad Rudrapatna"
date: 'December 16, 2022'
output: 
  pdf_document    
    number_sections: yes
header-includes:
- \usepackage{float}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(naniar)
library(UpSetR)

# wildlife strikes, passwords, freedom, indoor pollution, layoffs (kaggle), company reputations

# wealth and income, NFL/soccer attendance - results

# global plastic waste
```

```{r}
coast_vs_waste <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/coastal-population-vs-mismanaged-plastic.csv")

mismanaged_vs_gdp <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-mismanaged-plastic-waste-vs-gdp-per-capita.csv")

waste_vs_gdp <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-plastic-waste-vs-gdp-per-capita.csv") %>%
  select(Entity, Code, Year, `Per capita plastic waste (kilograms per person per day)`, `Total population (Gapminder)`)

plastic <- left_join(mismanaged_vs_gdp, waste_vs_gdp, by = c('Entity', 'Code', 'Year', 'Total population (Gapminder)')) %>%
  rename("Country" = "Entity",
         "Population" = "Total population (Gapminder)",
         "Per Capita GDP" = "GDP per capita, PPP (constant 2011 international $) (Rate)",
         "Per Capita Mismanaged Waste (Per Day)" = "Per capita mismanaged plastic waste (kilograms per person per day)",
         "Per Capita Waste (Per Day)" = "Per capita plastic waste (kilograms per person per day)") %>%
  filter(Year == '2010') %>%
  select(-Code)
```

```{r}
vis_miss(plastic, warn_large_data = F)
gg_miss_upset(plastic)
```

# conduct a complete case analysis
- response variable is created from mismanaged and total waste
- per capita GDP is the covariate of interest
- control for population

-> exclude countries without full information
-> only consider 2010 since that's when plastic waste data is available
-> GDP (based on 2011 international rates)

-> create response variable: quotient of Per Capita Mismanaged Waste (Per Day) / Per Capita Waste (Per Day)

```{r}
plastic2 <- plastic %>%
  na.omit() %>%
  select(Country, Population, `Per Capita GDP`, `Per Capita Mismanaged Waste (Per Day)`, `Per Capita Waste (Per Day)`) %>%
  mutate(`Mismanaged Waste Rate` = `Per Capita Mismanaged Waste (Per Day)` / `Per Capita Waste (Per Day)`)
```


Research Question: How is a country's per capita GDP associated with its rate of mismanaged plastic waste?

response variable: rate
beta regression?

```{r}
library(betareg)
m1 <- betareg(`Mismanaged Waste Rate` ~ `Per Capita GDP` + Population, data = plastic2)

summary(m1)
```



